Sketching for Density Estimation

### 2018-11-27 density.tex

\begin{lemma} \label{lem:compactor}
Let $k$ be an integer and assume we have black box access to a solver that for any $k$ inputs $x_1,\ldots,x_k$ obtains $k$ signs $s_1,\ldots,s_k$ such that
$$\max_q \left| \sum_i s_i f(x_i, q)\right| \leq \rho k$$
Then there exist a streaming algorithm requiring a memory buffer of $k$ input items that given a stream of length $n$ outputs a stream $z_1,\ldots,z_m$ with the following properties
\begin{itemize}
\item $\{z_i\}_i$ is a subset of $\{x_i\}$
\item $\E[m] = n/2$
\item For any fixed query $q$, the error associated with $q$, defined as
$$\text{Err}(q) = 2\sum_{i=1}^m f(z_i,q) - \sum_{i=1}^n f(x_i,q)  $$
Can be decomposed as a sum
$$\text{Err}(q) = \sum_{j=1}^{n/k} \text{Err}_j(q)$$
where the different $\text{Err}_j(q)$ are independent, have mean 0 and $|\text{Err}_j(q)| \leq \rho k$
\end{itemize}
The last property ensures that deterministically $|\text{Err}(q)| \leq \rho n$, and that w.h.p (via Chernoff's inequality) for a fixed $q$ 
$$|\text{Err}(q)| \lesssim \rho k\sqrt{n/k} = \rho \sqrt{kn}$$
\end{lemma}
\begin{proof}
The algorithm operates as follows. It keeps a buffer of $k$ items. Once the buffer fills with $x_1,\ldots,x_k$ it obtains the signs guaranteeing
$$\max_q \left| \sum_i s_i f(x_i, q)\right| \leq \rho k$$
Now, consider the set $P=\{i:s_i=1\}$ and $N=\{i:s_i=-1\}$.
$$ \sum_i s_i f(x_i, q) = \sum_i s_i f(x_i, q) + \sum_i f(x_i,q) - \sum_i f(x_i,q) = 2\sum_{i \in P} f(x_i,q) - \sum_i f(x_i, q)   $$
$$ -\sum_i s_i f(x_i, q) = \sum_i -s_i f(x_i, q) + \sum_i f(x_i,q) - \sum_i f(x_i,q) = 2\sum_{i \in N} f(x_i,q) - \sum_i f(x_i, q)   $$
Thus, we draw a random coin and either output to the stream the data points of $P$ or $N$. We call this operation a single compaction. Since $P,N$ form a partition of $[k]$ we get that the expected output length is $k/2$ for every $k$ inputs, insuring $\E[m]=n/2$. As for the error term w.r.t a fixed query $q$, it is clearly the sum of errors accumulated in each compaction. The error incurred in a compaction is either 
$\sum_i s_i f(x_i, q)$ or $-\sum_i s_i f(x_i, q)$ with equal probability. It follows that it is a Bernoulli r.v. with magnitude of at most $\rho k$ as required.
\end{proof}

@@

To achieve a deterministic bound we set all $k_h=k$, and set the compactors to choose the set $P$ or $N$ with the least cardinality, ensuring $n_h \leq n_{h-1}/2$ and $H \leq \log_2(n/k)$, thus
$$ \text{Err}(q) \leq \rho(k) \log_2(n/k) n $$

@@

Hence, if we aim for an error of $\eps n$ w.p.\ $1-\delta$ we set $k$ large enough such that $\rho(k) \leq c \eps / \log\log(1/\delta)$. Denoting by $\rho^{\dagger}$ the inverse function to $\rho$ this leads to a total memory requirement of 
$$ \rho^{\dagger}(c \eps / \log\log(1/\delta)) \log\log(1/\delta) $$
%=======
%$O(\log^3(\log(1/\delta))/\eps^2)$ whose error is at most $\eps n$ and failure probability at most $\delta$.
%\end{fact}
%\begin{proof}
%For simplicity, assume that $n$, the length of the stream, is known in advance. This assumption will be removed later.
%We begin with an obvious observation that sampling is always a possible first step if the stream is long enough.
%Consider sampling stream items with probably $p = \log(1/\delta)/n\eps^2$ and weighting the sampled items by $w = 1/p$.
%This gives a an $\eps n$ approximation for any single query $q$ with probability at least $1-\delta$. 
%Let $Q_\eps$ denote an epsilon-net over $f$. Specifically $\max_q \min_{q' \in Q_\eps} |F(q) - F(q')| \le \eps n$.
%Setting $\delta \le 1/|Q_\eps|$ will achieve a strong corset for using the union bound. 
%Note that $|Q_\eps|$ is usually exponential in the parameters of the problem and $\log(1/\delta)$ is polynomial in those parameters. 
%
%\noindent From here on, we assume $n = O(\log(1/\delta)/\eps^2)$ for some appropriately small value $\delta$.
%Let us begin with the simpler case where $\Delta_k = O(\sqrt{k})$. 
%Here, we will create $H$ buffers $B_1,...,B_H$ of size $k$ feeding one into each other. 
%The number items seen by buffer $h$ is $n_h = n/2^{h-1}$ and their weight is $w_h = 2^{h-1}$.
%The error induce by buffer $h$ is at most $w_h\frac{n_h}{k} \sqrt{k} = n/\sqrt{k}$. The total error is therefore at most $Hn/\sqrt{k}$.
%In order for the last buffer to not overflow we need to make sure that $n_H \le k$. Setting $H = \log(n/k)+1$ achieves that.
%Demanding that $Hn/\sqrt{k} \le \eps n$ and setting $k = (\log\log(1/\delta) + 1)^2/\eps^2$ we get that the total error is less than $\eps n$.
%Substituting we get that $H \le \log\log(1/\delta) + 1$. Finally, the total storage of the sketch is $Hk = O(\log\log(1/\delta) + 1)^3/\eps^2$.
% 
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(1)$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps)$ whose error is at most $\eps n$.
%\end{fact}
%\begin{proof}
%Proving this is more involved. It follows the same steps as the recent breakthrough result about streaming quantiles \cite{DBLP:conf/focs/KarninLL16} which exhibit $\Delta_k = O(1)$. The following proof can be thought of as a generalization of the proof in \cite{DBLP:conf/focs/KarninLL16}. 
%\end{proof}
%
%
%
%
%
%>>>>>>> 3fb6fdab63baebeec7715bcb29d19665e3023a8c

@@

\zk{Selecting $o(1/\eps^2)$ is claimed to be hard by \cite{DBLP:conf/focs/DeshpandeR10} and \cite{DBLP:conf/soda/DeshpandeRVW06}. We need to see if these hold in our setting or only hold for sampling.  Also, \cite{DBLP:conf/soda/GhashamiP14} showed that the impossibility result is limited to frequent directions like settings. Their assumptions don't hold here.}

>>>>

\begin{lemma} \label{lem:compactor}
Let $k$ be an integer and assume we have black box access to a solver that for any $k$ inputs $x_1,\ldots,x_k$ obtains $k$ signs $s_1,\ldots,s_k$ such that
$$\max_q \left| \sum_i s_i f(x_i, q)\right| \leq \rho k$$
Then there exist a streaming algorithm requiring a memory buffer of $k$ input items that given a stream of length $n$ outputs a stream $z_1,\ldots,z_m$ with the following properties
\begin{itemize}
\item $\{z_i\}_i$ is a subset of $\{x_i\}$
\item $\E[m] = n/2$
\item For any fixed query $q$, the error associated with $q$, defined as
$$\text{Err}(q) = 2\sum_{i=1}^m f(z_i,q) - \sum_{i=1}^n f(x_i,q)  $$
Can be decomposed as a sum
$$\text{Err}(q) = \sum_{j=1}^{n/k} \text{Err}_j(q)$$
where the different $\text{Err}_j(q)$ are independent, have mean 0 and $|\text{Err}_j(q)| \leq \rho k$
\end{itemize}
The last property ensures that deterministically $|\text{Err}(q)| \leq \rho n$, and that w.h.p.\ (via Chernoff's inequality) for a fixed $q$ 
$$|\text{Err}(q)| \lesssim \rho k\sqrt{n/k} = \rho \sqrt{kn}$$
\end{lemma}
\begin{proof}
The algorithm operates as follows. It keeps a buffer of $k$ items. Once the buffer fills with $x_1,\ldots,x_k$ it obtains the signs guaranteeing
$$\max_q \left| \sum_i s_i f(x_i, q)\right| \leq \rho k$$
Then, the algorithm output to the stream the with probability $1/2$ each data points $\{x_i \; | \; s_i = 1\}$ or $\{x_i \; | \; s_i = -1\}$ with twice the weight.
$$\sum_{i ,\; s_i=1} 2f(x_i, q) = \sum_{i} f(x_i, q) +  \sum_{i} s_i f(x_i, q)$$
$$\sum_{i ,\; s_i=-1} 2f(x_i, q) = \sum_{i} f(x_i, q) - \sum_{i} s_i f(x_i, q)$$

%Now, consider the set $P=\{i:s_i=1\}$ and $N=\{i:s_i=-1\}$.
%$$ \sum_i s_i f(x_i, q) = \sum_i s_i f(x_i, q) + \sum_i f(x_i,q) - \sum_i f(x_i,q) = 2\sum_{i \in P} f(x_i,q) - \sum_i f(x_i, q)   $$
%$$ -\sum_i s_i f(x_i, q) = \sum_i -s_i f(x_i, q) + \sum_i f(x_i,q) - \sum_i f(x_i,q) = 2\sum_{i \in N} f(x_i,q) - \sum_i f(x_i, q)   $$
%Thus, we draw a random coin and either output to the stream the data points of $P$ or $N$. We call this operation a single compaction. 
%\el{this is clunky way to say this...}
The expected output length is $k/2$ for every $k$ inputs, insuring $\E[m]=n/2$. 
As for the error term w.r.t.\ a fixed query $q$, it is clearly the sum of errors accumulated in each compaction. The error incurred in a compaction is either 
$\sum_i s_i f(x_i, q)$ or $-\sum_i s_i f(x_i, q)$ with equal probability. It follows that it is a Bernoulli r.v. with magnitude of at most $\rho k$ as required.
\end{proof}

@@

To achieve a deterministic bound we set all $k_h=k$, and set the compactors to choose the signed set with the least cardinality, ensuring $n_h \leq n_{h-1}/2$ and $H \leq \log_2(n/k)$, thus
$$ \text{Err}(q) \leq \rho(k) \log_2(n/k) n $$

@@

Hence, if we aim for an error of $\eps n$ w.p.\ $1-\delta$ we set $k$ large enough such that $\rho(k) \leq c \eps / \log\log(1/\delta)$. Denoting by $\rho^{\dagger}$ the inverse function to $\rho$ this leads to a total memory requirement of 
$$ \rho^{\dagger}(c \eps / \log\log(1/\delta)) \log\log(1/\delta) $$

@@

\el{Selecting $o(1/\eps^2)$ is claimed to be hard by \cite{DBLP:conf/focs/DeshpandeR10} and \cite{DBLP:conf/soda/DeshpandeRVW06}. We need to see if these hold in our setting or only hold for sampling.  Also, \cite{DBLP:conf/soda/GhashamiP14} showed that the impossibility result is limited to frequent directions like settings. Their assumptions don't hold here.} \zk{We can get $1/\eps^2$ though which is better than $\sqrt{d}/\eps$ as long as $\eps > 1/\sqrt{d}$. If $\eps < 1/\sqrt{d}$ then $\sqrt{d}/\eps > d$ - shouldn't it be trivial to get a coreset of size $d$?}
